```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = FALSE, 
  message=FALSE, 
  warning=FALSE, 
  fig.width = 10,
  fig.align = "center"
  )

library( here )      # for the directory
library( ggraph )    # for plotting
library( igraph )    # for working with graphs
library( ggplot2 )   # for plotting
library( gridExtra ) # for plotting multiple plots
library( grid )      # to include the null graphical object
library( reshape2 )  # for reshaping the matrix data for a plot

```

# Network Description {.unnumbered}

Understanding the basic structure of a network is the first step toward making informed, data-driven decisions in crime analysis. Networks, whether simple or complex, contain key features that offer valuable insights into relationships, patterns, and points of vulnerability. As crime analysts, the ability to describe these features (such as the number of connections), the centrality of nodes, and the overall shape of the network, is essential for identifying trends and prioritizing interventions. But how do we translate raw data into these insights? Before we get too deep into analysis, we should start by doing some basic description of a network to get our hands dirty.

By the end of the chapter, you should be able to answer this question:

-   How can I describe some simple features of a network?

```{r, graphs-define}

# ----
# This creates the graphs for the plots below

# Create the graph objects
ugraph  <- graph_from_data_frame( 
  data.frame( 
    from = c( 1, 2, 3, 3, 4 ),
    to   = c( 2, 3, 4, 5, 5 ) ), 
  directed = FALSE )

digraph  <- graph_from_data_frame( 
  data.frame( 
    from = c( 1, 2, 3, 3, 4, 4, 5, 5  ),
    to   = c( 2, 3, 4, 5, 3, 5, 3, 4  ) ), 
  directed = TRUE )

# Set the random seed to render the same plot
set.seed( 507 )

# Set a fixed layout using the Fruchterman-Reingold layout
layout <- layout_with_fr( ugraph )
dilayout <- layout_with_fr( digraph )

# Set the labels
custom_labels <- c( "Jen","Tom","Bob","Leaf","Jim" )

# Assign the labels to the graph nodes
V( ugraph )$name <- custom_labels
V( digraph )$name <- custom_labels

```

```{r, plots-define}

# create the undirected graph
u_graph <-
  ggraph( ugraph, 
          layout = layout ) +                  
  geom_edge_link( color = "black", width = 0.8 ) +  
  geom_node_point( color = "#28a88d", size = 15 ) +
  geom_node_text( aes( label = name ), 
                 color = "black", 
                 size = 5,        
                 vjust = 0.5,     
                 hjust = 0.5 ) +
  ggtitle( "Undirected Graph" ) + 
  scale_x_continuous( expand = expansion( mult = c( 0.2, 0.2 ) ) ) +
  scale_y_continuous( expand = expansion( mult = c( 0.2, 0.2 ) ) ) +
theme_void() 

# create the directed graph
d_graph <- 
  ggraph( digraph, 
          layout = dilayout ) +                  
  geom_node_point( color = "#fc23fc", size = 15 ) +
  geom_node_text( aes( label = name ), 
                 color = "black", 
                 size = 5,        
                 vjust = 0.5,     
                 hjust = 0.5) +
  geom_edge_link( aes( start_cap = label_rect( node1.name ), 
                     end_cap = circle( 5, 'mm' ) ),
                 arrow = arrow( length = unit( 0.02, "npc" ) ), 
                 color = "black", width = 0.8) +
  ggtitle( "Directed Graph" ) + 
  scale_x_continuous( expand = expansion( mult = c( 0.2, 0.2 ) ) ) +
  scale_y_continuous( expand = expansion( mult = c( 0.2, 0.2 ) ) ) +
  theme_void()  

```

## Simple Network Summary

So you have your network data, now what? The next step is to summarize its most critical features, which can guide your analysis. First, consider the size: How *big* is the network? This is the number of nodes involved, giving a sense of its scale. Next, think about how tightly connected the network is---this is the *density*, or the proportion of actual connections relative to all possible ones. Another key question is separation: How fragmented or divided is the network? Understanding its *components* or *subgraphs* can highlight isolated clusters. Finally, consider the network's compactness, or its *diameter*, how far do nodes have to travel through connections to reach others? These four basic measures lay the groundwork for describing and understanding the structure of any network you encounter in your analysis.

### Size

Put simply: how big is it? The size of the network is just the number of nodes in the network. We represent this with $g$.

In the graphs below, what is $g$?

```{r}

# Display the plots side by side
grid.arrange( u_graph, nullGrob(), d_graph, ncol = 3, widths = c( 1, 0.2, 1 ) )

```

### Density

Of the ties that could exist, what fraction are observed? This is the *density* of a graph: the number of edges compared to how many ties there could be.

#### Undirected Graph

The density of an **undirected** graph is given by: $$\frac{2L}{g(g-1)}$$ where $L$ is the number of edges and $g$ is the number of nodes. So, if a network has 5 nodes (i.e. $g$ is 5) then we could observe $g(g-1) = 5*(5-1) = 5*4 = 20$ edges. Take a look at the undirected graph below. What is the density?

```{r}

u_graph

```

First, we identify the quantities: $L$ is 5 and $g$ is 5. Now, let's just plug through the math: $$\frac{2L}{g(g-1)} = \frac{2*5}{5*(5-1)} = \frac{10}{20} = 0.5$$ So, 0.5, or half, of the ties that could occur in this network did occur.

#### Directed Graph

The density of a **directed** graph is given by: $$\frac{L}{g(g-1)}$$ where $L$ is the number of edges and $g$ is the number of nodes. So, if a network has 5 nodes (i.e. $g$ is 5) then we could observe $g(g-1) = 5*(5-1) = 5*4 = 20$ edges. Take a look at the directed graph below. What is the density?

```{r}

d_graph

```

First, we identify the quantities: $L$ is 8 and $g$ is 5. Why is $L$ 8? Since the graph is directed, a dyad (or two nodes) can have two ties. For example, there is a tie from Jim to Leaf and a tie from Leaf to Jim. If we count up all the arrowheads, we get to 8. Now, let's just plug through the math: $$\frac{L}{g(g-1)} = \frac{8}{5*(5-1)} = \frac{8}{20} = 0.4$$ So, 0.4 of the ties that could occur in this network did occur.

### Components

So far, we have seen graphs where the nodes are connected either directly or indirectly. For example, in the directed graph above, Jen is not directly connected to every node, but you can trace a path from Jen to all other nodes in the graph. But what about graphs where the nodes are not able to reach other nodes? In other words, what about graphs where there are separated groups of nodes? These disconnected **subgraphs** are referred to as *components*.

Take a look at the plot below. How many components are there?

```{r}

# set the seed to reproduce the plot
set.seed( 507 )

# create the edgelist
edges <- data.frame( from = sample(1:150, 100, replace = TRUE ),
                     to = sample(1:150, 100, replace = TRUE ) )

# remove duplicate edges
edges <- unique( edges )

# remove self-loops
edges <- edges[ edges$from != edges$to, ]

# create the graph
mod_graph  <- graph_from_data_frame( edges, directed = FALSE ) 

# identify the components
components <- components( mod_graph )

# find the largest component
largest_component <- which.max( components$csize )

# add component membership as a vertex attribute
V( mod_graph )$component <- components$membership

# add labels only for the nodes in the largest component
V( mod_graph )$largest <- ifelse( V( mod_graph )$component == largest_component, V( mod_graph )$component, NA )

# set the layout
mod_layout <- layout_with_fr( mod_graph )

# plot the network
comp_graph <-
  ggraph( mod_graph, 
          layout = mod_layout ) +                  
  geom_edge_link( ) +  
  geom_node_point( aes( color = as.factor( component ) ), size = 5, show.legend = FALSE ) + 
  geom_node_text( aes( label = largest ), color = "black" ) + 
  labs(color = "Component") +
  ggtitle( "Undirected Graph w/ Components" ) + 
  theme_void() 

comp_graph

```

To figure it out, just count up the number of separate subgraphs (or different colors of nodes). In this plot, there are 15 components. Note that the components are not all the same. Some are larger than others. In a graph that has different-sized components, the *first component* is the subgraph (i.e., component) with the most nodes. In this plot, the nodes in the first component are labeled "1". The *second component* of a graph is the component that has the second-most nodes, and the *third component* has the third-most nodes (and so on). In the plot, what are the *second* and *third* components?

The components of a graph are a useful tool for helping us think conceptually about what our network means. For example, @papachristos2014network conducted a study of homicide, examining how individuals may be disproportionately exposed to violence. Violence often spreads between components through a small number of brokers who connect otherwise isolated factions. Identifying and targeting these brokers or the most violent components can significantly reduce retaliatory violence. Isolated components also offer opportunities for more focused community interventions. By understanding these network structures, law enforcement can prevent the spread of violence and effectively target interventions.

### Diameter

A final question we can ask is how *compact* is the network? In other words, can nodes reach other nodes directly and indirectly fairly easily? Or, are nodes really far from each other?

To understand diameter, let's set up a definition: A *step* is movement along an edge. Let's think about how this works. For example, take two nodes $A$ and $B$ who are directly connected. How many steps is $A$ from $B$? Since $A$ is directly tied to $B$, the number of steps is 1. Now, take another example where $A$ and $B$ are *not* directly connected. Rather, suppose $A$ is connected to $D$, whom is connected to $C$, whom is connected to $B$, like this: $A-D-C-B$. How many steps is $A$ from $B$? Well, how many edges are there between them? There are 3 edges.

We can extend this to an entire graph by calculating the steps for every pair of nodes (i.e., dyads) in the graph. When we do this, we can define two other terms:

-   A *shortest path* is the fewest number of steps node $A$ has to take to reach node $B$;

-   The *diameter* is the longest of **all** the shortest paths in the network.

#### Undirected Graph

Let's see how this works in an **undirected** graph. In the plot below, what is the shortest path from Leaf to Jen?

```{r}

u_graph

```

To answer this question, we need to first figure out what the paths are, then figure out which is the shortest (i.e., has the fewest edges to traverse). In this plot, there are two:

-   $Leaf-Jim-Bob-Tom-Jen$
-   $Leaf-Bob-Tom-Jen$

The second path, $Leaf-Bob-Tom-Jen$, is the shortest path.

Now, what is the *diameter* of the network? To figure this out, we would identify the shortest paths for all pairs of nodes, then see which is the largest. So, taking a look at this plot, what is the *diameter*? It is 3. The longest shortest paths (I know, it sounds confusing) are $Leaf-Bob-Tom-Jen$ and $Jim-Bob-Tom-Jen$. Both are of length 3.

#### Directed Graph

Paths with *directed graphs* can be a bit tricky to comprehend due to the directionality of the ties. For example, how many steps is $A$ from $B$ in this graph, $B \rightarrow D \rightarrow C \rightarrow A$? If we focus on the direction of the ties, then $A$ cannot reach $B$. But, if we **ignore** the direction of the ties, then $A$ is 3 steps from $B$. Often, we ignore the directionality of a network to calculate the diameter. But, in some cases, it might not make sense to do so. Just keep this in mind as you are working with different networks as a crime analyst.

<br>

## Test your Knowledge

-   What are the key features of a network that crime analysts can focus on for identifying trends and prioritizing interventions?
-   How do you calculate the size of a network, and why is it important to know?
-   What is network density, and how is the calculation different for undirected and directed graphs?
-   Why is it important to examine the components of a network?
-   How do you identify the largest component in a graph, and why might this component be particularly important?
-   What is the diameter of a network, and why is it a significant measure of network compactness?

## Summary

This chapter focused on how crime analysts can gain an understanding of a network by describing some basic structures. We focused on key network features (size, density, components, and diameter) to gain insights into relationships and patterns. Examining these properties can help analysts identify areas for targeted interventions. In the next chapters, we will begin focusing on a key idea in social network analysis: *centrality*.
